---
title: "Untitled"
format: html
---

```{r}
library(tidyverse)
library(janitor)

session_df <- read_csv("full_session_data.csv")
survey_df <- read_csv("survey_data.csv")

```

```{r}

```


## Task Performance Scoring
Participants completed two binary subtasks: suspect identification and location zone identification (coded zone_correct = 1 if “Loading bay”). A combined accuracy score (total_correct = suspect_correct + zone_correct; range 0–2) was computed. A proportional score can be defined as prop_correct = total_correct / 2. (If additional components—status, building, floor—are intended but not yet coded into total_correct, adjust the formula and descriptive text accordingly.)

## Descriptive Statistics
Group means in total correct answers (CONTROL vs EXPERIMENTAL): 3.12 vs 2.92 (these appear to assume a 5‑item scale; reconcile with the current 0–2 construction).  
By native English status: Non‑native (n = 15) M = 2.6; Native (n = 6) M = 4.0 (again, verify scale alignment).

(Recommend regenerating descriptives after fixing total_correct definition.)

## Group Difference: Native English
Welch’s t-test: t(8.34) = –2.36, p = 0.044; Hedges’ g = –1.06, 95% CI [–2.05, –0.03]. Sign indicates ordering (first level has lower mean). Large absolute effect; wide interval and small, unbalanced sample reduce precision. Bayesian test: BF10 = 2.94 (anecdotal to moderate evidence).

## Multiple Regression
Model: total_correct ~ group + native_english + duration_mins + n_frust + n_neg + n_engaged + n_silence  
R² = 0.53, Adjusted R² = 0.27, F(7,13) = 2.07, p = 0.122.

| Predictor        | Estimate | SE      | t      | p       |
|------------------|---------:|--------:|-------:|--------:|
| (Intercept)      | 4.176    | 0.727   | 5.743  | < .001  |
| group (EXP)      | -0.760   | 0.568   | -1.338 | 0.204   |
| native_english1  | -0.666   | 0.329   | -2.021 | 0.064   |
| duration_mins    | 0.030    | 0.071   | 0.419  | 0.682   |
| n_frust          | -0.816   | 0.373   | -2.186 | 0.048   |
| n_neg            | 0.042    | 0.308   | 0.135  | 0.894   |
| n_engaged        | 0.169    | 0.174   | 0.970  | 0.350   |
| n_silence        | -0.135   | 0.133   | -1.020 | 0.327   |

Interpretation: Higher frustration count is associated with lower accuracy (p < .05). Native English shows a trend (p = .064). Experimental vs Control difference is not statistically reliable. Other session metrics show no association. Given small n (21) and many predictors (7), potential overfitting and unstable estimates are concerns; consider a reduced model or penalized approach.

## Multiple Regression (Reduced Model)

A reduced linear model examined group (Experimental vs Control) and native English status as predictors of total task accuracy. The model explained 29.8% of the variance (R² = 0.30; adjusted R² = 0.22), F(2, 18) = 3.81, p = 0.042.

| Predictor              | Estimate | SE     | t      | p      |
|------------------------|---------:|-------:|-------:|-------:|
| (Intercept)            | 3.72     | 0.46   | 8.06   | < .001 |
| group (EXPERIMENTAL)   | -0.61    | 0.54   | -1.14  | 0.270  |
| native_english1        | -0.79    | 0.29   | -2.73  | 0.014  |

Interpretation: Native English participants scored higher (coefficient negative due to sum contrast coding: level “1” sits above the grand mean). The Experimental group did not differ reliably from Control (p = 0.27). The effect associated with native English status is moderate to large given the scale range, but precision is limited by small and unbalanced sample sizes.

Report: “In the reduced model, native English status predicted higher accuracy (β = -0.79, SE = 0.29, p = .014), whereas Experimental vs Control condition did not (β = -0.61, SE = 0.54, p = .27), R² = .30.”

(If scale definition for total_correct is 0–2 rather than 0–5, reconcile before final reporting; current intercept suggests a wider scale.)

## Cautions
- Scale definition inconsistency (reported means exceed possible range if total_correct truly maxes at 2). Recompute total_correct or revise descriptive section.
- Small, unbalanced groups inflate standard errors; effect sizes should be emphasized with CIs.
- Potential multicollinearity among affect counts (n_frust, n_neg, n_engaged, n_silence); inspect variance inflation factors before final reporting.

## Summary
Evidence suggests a performance difference by native English status and a negative association with frustration, but intervention group effects were not supported. Results should be interpreted cautiously due to sample size and measurement inconsistencies.

## Example Reporting Sentence
“Frustration count was a significant negative predictor of task accuracy (β = -0.82, p = 0.048), while native English status showed a trend (p = 0.064); no effect of experimental condition was observed (p = 0.20). Overall model fit was moderate (R² = 0.53), but estimates are imprecise given n = 21.”
